OptiFit: a fast method for fitting amplicon sequences to existing OTUs

### 2021-06-09 paper/paper.Rmd
Separate intro paragraphs for _de novo_ and reference clustering

There are three main categories of OTU clustering algorithms: closed reference,
open reference, and _de novo_ clustering. 
Closed reference methods assign sequences to a set of pre-made OTUs generated
from reference sequences.
If a query sequence is not within the distance threshold to any of the reference
sequences, it is discarded.
While reference-based clustering is generally fast, it is limited by the
diversity of the reference database.
Rare or novel sequences in the sample will be lost if they are not represented
by a similar sequence in the database.
_De novo_ methods cluster sequences based on their distance to each other,
without the use of an external reference.
_De novo_ clustering overcomes the limitations of reference databases by
considering only sequences in the dataset, but is more computationally intensive
and generates different OTU assignments when new sequences are introduced.
Unstable OTU assignments make it difficult to use _de novo_ clustering to
compare taxonomic composition of communities between studies or to use machine
learning models trained with _de novo_ OTUs to make predictions on new data.
Open reference methods take a hybrid approach, first performing closed reference
clustering, then any sequences that cannot be assigned to reference OTUs are
clustered _de novo_ to create additional OTUs.
Previous studies found that the OptiClust _de novo_ clustering algorithm created
the highest quality OTU assignments of all clustering methods based on the
Matthews correlation coefficient (MCC) [@westcott_opticlust_2017].
As a result, we have recommended OptiClust as the preferred method for OTU
clustering whenever OTU stability is not required.

- **TODO: current method for open/closed is vsearch against greengenes.**
- **TODO: use word "map" for what vsearch does, "fit" for what optifit does.**
- **TODO**: 2 categories of clustering: _de novo_ and reference based. separate paragraphs. describe opticlust first in de novo paragraph. 2nd paragraph: ref methods are good cause they're fast and don't use much ram. dependent on order of db. people use greengenes, which are rep seqs from 3% otus from full length.
- reader should know what opticlust is, closed & open ref clustering is, strengths & weakness are of each. then we solve these problems.
- **TODO:** note that greengenes is defunct now?!

To overcome the limitations of _de novo_ clustering while maintaining OTU
quality, we developed OptiFit, a reference-based clustering algorithm in the
mothur program which takes existing OTUs as the reference to fit new sequences
to. 
**TODO: more words here?** 
Here, we tested the OptiFit algorithm with the reference as a database or _de
novo_ OTUs and compared the performance to existing tools.
To evaluate the OptiFit algorithm and compare to existing methods, we used four
published datasets isolated from soil [@johnston_metagenomics_2016], marine
[@henson_artificial_2016], mouse gut [@schloss_stabilization_2012], and human
gut [@baxter_microbiota-based_2016] samples.

@@

When fitting to public databases, OTU quality dropped in closed reference mode
to different degrees depending on the database and dataset source, and no more
than half of query sequences were able to be fit to OTUs across any
dataset/database combination. 
This may reflect limitations of reference
databases, which are unlikely to contain sequences from rare and novel microbes.
This drop in quality was most notable with RDP, which contains only about 21,000
sequences compared to over 200,000 sequences in SILVA and Greengenes each at the
time of this writing.
We recommend that users who require an independent reference database opt for
large databases with good coverage of microbial diversity.
Since OptiClust performs faster than open reference OptiFit and creates higher
quality OTUs than closed reference OptiFit with the database strategy, we
recommend using OptiClust rather than fitting to a database whenever stable OTUs
are not required for the study at hand.

>>>>

There are two main categories of OTU clustering algorithms: _de novo_ and
reference-based.
OptiClust, the _de novo_ clustering method provided by the mothur software
program, uses the distance score between all pairs of sequences in the dataset
to cluster them into OTUs by maximizing the Matthews correlation coefficient
(MCC) [@westcott_opticlust_2017].
This approach takes into account the distances between all pairs of sequences
when assigning query sequences to OTUs, in contrast to other tools such as
VSEARCH which only considers the distance between the query sequence and a
representative centroid sequence in the OTU.
A limitation of _de novo_ clustering is that different OTU assignments will be
produced when new sequences are added to a dataset, making it difficult to use
_de novo_ clustering to compare OTUs between different studies.
Furthermore, since _de novo_ clustering requires distance calculations between all
sequences in a dataset, the execution time can be slow for very large datasets.

Reference clustering attempts to overcome the limitations of _de novo_
clustering methods by using a representative set of sequences from a database,
with each reference sequence seeding an OTU.
Commonly, the Greengenes set of representative full length sequences clustered
at 97% similarity is used as the reference with VSEARCH
[@desantis_greengenes_2006].
Query sequences are then assigned to OTUs based on their similarity to the
reference sequences.
Any query sequences that are not within the distance threshold to any of the
reference sequences are either thrown out (closed reference mode) or clustered
_de novo_ to create additional OTUs (open reference mode).
While reference-based clustering is generally fast, it is limited by the
diversity of the reference database.
Rare or novel sequences in the sample will be lost in closed reference mode if
they are not represented by a similar sequence in the database.
Previous studies found that the OptiClust _de novo_ clustering algorithm created
the highest quality OTU assignments of all clustering methods based on the
Matthews correlation coefficient (MCC) [@westcott_opticlust_2017].
As a result, we have recommended OptiClust as the preferred method for OTU
clustering whenever OTU stability is not required.

To overcome the limitations of _de novo_ clustering while maintaining OTU
quality, we developed OptiFit, a reference-based clustering algorithm in the
mothur program which takes existing OTUs as the reference to fit new sequences
to. 
In contrast to other tools, OptiFit considers all pairwise distance scores
between reference and query sequences when assigning sequences to OTUs in order
to produce OTUs of the highest possible quality.
Here, we tested the OptiFit algorithm with the reference as a database or _de
novo_ OTUs and compared the performance to existing tools.
To evaluate the OptiFit algorithm and compare to existing methods, we used four
published datasets isolated from soil [@johnston_metagenomics_2016], marine
[@henson_artificial_2016], mouse gut [@schloss_stabilization_2012], and human
gut [@baxter_microbiota-based_2016] samples.

@@

When fitting to public databases, OTU quality dropped in closed reference mode
to different degrees depending on the database and dataset source, and no more
than half of query sequences were able to be fit to OTUs across any
dataset/database combination. 
This may reflect limitations of reference
databases, which are unlikely to contain sequences from rare and novel microbes.
This drop in quality was most notable with RDP, which contains only about 21,000
sequences compared to over 200,000 sequences in SILVA and Greengenes each at the
time of this writing.
**TODO:** note that greengenes is defunct now?!
We recommend that users who require an independent reference database opt for
large databases with good coverage of microbial diversity.
Since OptiClust performs faster than open reference OptiFit and creates higher
quality OTUs than closed reference OptiFit with the database strategy, we
recommend using OptiClust rather than fitting to a database whenever stable OTUs
are not required for the study at hand.
